Insecticide resistance evolution with mixtures and sequences : a model-based explanation.

### 2018-01-02 inst/documents/paper2/paper2_resistance_mechanisms_mixtures.Rmd
renumber figure references in text

```{r, eval=TRUE, include=FALSE}
#!!!! PAPER PROTOCOL
#!!!! submission MS, knit to word_document 
#!!!!  in WORD: delete figs, set double space, add line numbers 
#!!!!  (spacing & line numbers work for pdf output but not for word)
#!!!! submission figs, set this chunk to eval=TRUE
#!!!! previous drafts knitted to pdf with pdf_document in YAML and this chunk eval=FALSE

# options to create final publication quality figures
# causes the document to fail, but the figs do get stored in pap2figs/
library(knitr)
opts_chunk$set(dev="tiff",
               dev.args=list(compression="lzw"),
               dpi=300,
               cache=FALSE,
               fig.path='pap2figs/')
```

@@

The calculation of fitness for each genotype when considering just a single insecticide is illustrated in Figure 1. In the model itself this is repeated for both insecticides and the fitnesses multiplied. Firstly the exposure input determines the proportion of the population in the left and right panels (exposed and not exposed). For those that are exposed (left panel) insecticide effectiveness sets the fitness for SS and resistance restoration 'restores' a portion of the fitness for RR. A value of 1 for resistance restoration would bring the fitness of RR back to 1, equivalent to the unexposed SS and meaning the RR genotype is completely unaffected by the insecticide. A value of 0 for resistance restoration would lead to the RR having the same fitness as SS in the presence of the insecticide and effectively there would be no resistance. We use 'Resistance restoration' to represent how much the RR genotype restores fitness in the presence of the insecticide. Cost of resistance determines the fitness of RR in the portion of the population not exposed to the insecticide.

Dominance of restoration determines how the fitness for SR sits between that of SS and RR for those that are exposed to the insecticide. For those that are not exposed, fitness of SS is set to 1 by definition and resistance cost determines the fitness of RR. Dominance of cost determines how the fitness for SR sits between that of SS and RR in the absence of the insecticide (Fig 1).

The simulation proceeds through non-overlapping generations. In each generation selection is represented by multiplying genotype frequencies in the population by their relative fitness, the latter depending on their absolute fitness (Fig 1) and level of exposure. This acts to make the fitter alleles more common over time. Separate sexes and standard sexual reproduction with recombination is included. There is also the potential to set exposure rates to be different for males and females to represent their different behaviours but we do not explore that parameter here.

@@

A major aim of this paper is to us modern techniques to make the modelling accessible to a wider community. The model is implemented in R [@RCoreTeam2016], the code is open-source and hosted on Github [@South2017] including the code to generate both the figures and text of this paper. We provide an on-line user interface [@South2017a] (Fig 2) accessing the same code and enabling readers with no coding experience to change inputs and run the model themselves.

@@

### Single insecticide
For single insecticide use, higher values of insecticide effectiveness, exposure, resistance restoration and its dominance all resulted in faster resistance spread (Fig 3 A-D) and thus fewer generations to reach a resistance threshold of 0.5. This makes intuitive sense as all increase the impact of the insecticide and thus increase selection for resistance. Similarly, higher values of starting resistance frequency also resulted in shorter times to resistance thresholds (Fig 4A) through the even simpler mechanism that a smaller cahnge in frequency is required to reach the threshold.

Cost of resistance (Fig 4B) and its dominance (not shown) had the opposite effect. Higher values led to a slower development of resistance. Again this is intuitive because costs reduce the advantage of resistance and decrease selection. These effects of inputs on single insecticide use are summarised in Table 2.

@@

### Two insecticides
When two insecticides are used we can compare the relative performance of concurrent use in a mixture with sequential use. Each panel in Figures 5-9 compares mixture and sequence for a single combination of inputs. To help understand the mechanisms influencing this relative performance we start with a single base scenario and demonstrate changing inputs individually. In this base scenario we set effectiveness, exposure, resistance restoration, and dominance at 0.5, the starting frequencies of resistance to 0.01 and cost of resistance to 0. Resistance arises slower for sequential than mixture for this base scenario (Fig 5A). Resistance to both insecticides in the mixture follow the same path and reach the threshold at the same time (as would be expected given that they have identical input parameters). When used in sequence the curve (dashed) for each insecticide individually is steeper than in the mixture but, because they happen one after the other, it takes longer for both to reach the resistance threshold (Fig 5A). Thus the time-to-resistance for the mixture divided by the sequence is less than 1 at 0.8.

#### Insecticide exposure and effectiveness
Increasing exposure (the proportion of insects that come into contact with the insecticides) (Fig 5C) decreases the time-to-resistance for both the sequential and mixture strategy. The effect on the mixture is greater so time-to-resistance remains longer for the sequence (the ratio decreasing from 0.8 to 0.6). In contrast keeping the exposure constant and increasing effectiveness (the proportion of SS insects that are killed by contact) of one of the insecticides (Fig 5B) results in a longer time-to-resistance for the mixture relative to the sequence (and the ratio increasing from 0.8 to 1.2). The longer time-to-resistance for the mixture results from a slower increase in resistance for the less effective insecticide. Resistance to the less effective insecticide in the mixture increases slowly initially and then speeds up after the more effective insecticide has reached the resistance threshold. This points a mechanism whereby the more effective insecticide initially slows the rate of evolution of resistance to the less effective one.

The result is that a more effective insecticide increases time-to-resistance when used in a mixture (compare the solid lines in Fig 5A & B). This is opposite to what happens when used in sequence (compare the red dashed line in Fig 5A & B) and in isolation (Fig 2A), when more effective insecticides shortened times-to-resistance.

In summary increasing effectiveness can favour mixtures moving from column 1 to 2 in Figure 5, and increasing exposure can favour sequences moving from row 1 to 2 in Figure 5. Reduced male exposure with respect to female, as might be expected for male mosquitoes not seeking blood meals, had a similar effect to changing overall exposure. Reducing male exposure (not shown) increased times-to-resistance and favoured mixtures over sequences.

Increasing the effectiveness further of either or both insecticides increases the performance of mixtures over sequences in terms of time-to-resistance. Figure 6A uses the same inputs as Fig 5B. From this scenario a higher effectiveness for insecticide2 (Fig 6C) or insecticide 1 (Fig 6B) or both (Fig 6D) all result in a greater positive difference in time-to-resistance for mixture over sequence. This again points to how more effective insecticides can slow resistance to another when used in a mixture.

#### Resistance restoration and dominance
Increasing resistance restoration or its dominance decreases time-to-resistance for both sequences and mixtures (Fig 7) in the same way that it did for sole use (Fig 3). The result is that increasing either or both of dominance and resistance restoration (Figs 7B,C,D) doesn't change the relative time-to-resistance for mixtures and sequences from that in the base scenario (Fig 7A) where it takes resistance longer to develop for the sequence. So, unlike effectiveness and exposure, the levels of resistance restoration and dominance do not effect whether a mixture or a sequence are likely to be favoured.

#### Starting frequencies of resistance
Changing the starting frequency of resistance had a similar effect on time to resistance for both sequences and mixtures and thus had little effect on their relative performance. For example taking the base scenario and reducing the starting frequency of one resistance allele did not change from sequence being favoured (compare Fig 8C to 8A). Similarly taking a scenario in which time-to-resistance is longer for a mixture (Fig 8B) and decreasing the starting frequency of that resistance allele did not change the fact that the mixture was favoured (Fig 8D).

#### Costs of resistance
Increasing the cost of resistance to one insecticide in a mixture leads resistance to that insecticide to evolve more slowly (compare the solid red line to the solid blue line in Figure 9C, the red line is for the resistance with the higher cost). The resistance with the higher cost also increases more slowly in the mixture relative to when it is used in sequence (compare the solid red line to the dotted red line in Fig 9C). The greater effect of cost on mixtures means the advantage of sequences over mixtures in Fig 9A (and Fig 5A) is removed in Fig 9C. With costs set to 0 time-to-resistance for mixture divided by sequence is 0.8 in Fig 9A and 1 in Fig 9C. At higher insecticide effectiveness the benefit of mixture over sequence is improved as seen by comparing Fig 9D to 9B.

@@

#### Linkage disequilibrium
It has previously been suggested [@Curtis1985][@WHO2012] that using an insecticide to which there was existing resistance in a mixture with a new insecticide with little resistance could lead to more rapid evolution of resistance to the latter through a process termed linkage disequilibrium. Figure 8 C&D show how our model predicts that linkage disequilibrium does not lead to more rapid evolution of resistance to the newer insecticide. In Figures 8 C&D the newer insecticide with lower starting resistance (in red) leads to faster resistance when used alone (red dotted) than when used in a mixture (red solid).

@@

When single insecticide use was investigated (Figs 3-4) higher values of parameters that increased the selective advantage of resistance all led to more rapid evolution resistance, whereas higher values for cost led to slower evolution of resistance (Fig 4B). Mechanisms explaining the evolution of resistance to single insecticide use, are summarised in Table 2. Referring to figure 1 can help explain these responses. Exposure and effectiveness act by decreasing the fitness of SS and SR more than RR, dominance of restoration acts by increasing fitness of SR and resistance-restoration acts by increasing fitness of RR & SR. Cost acts by decreasing the fitness of RR and SR among those mosquitoes  not exposed.

#### Two insecticides
Responses to input parameters are different when two insecticides are used either in sequence or in mixtures as compared to when used alone. These differences can help us to understand the dynamics of resistance evolution. In each panel of Figures 5-9 the solid lines, indicating that an insecticide is being used at the same time as another in a mixture, are always shallower than the same colour dashed line indicating the insecticide is being used in isolation as part of a sequence. The shallower curves for the mixture can be explained by each insecticide killing individuals that are resistant to the other insecticide. Because resistant individuals are being killed, the selection pressure for that resistance is lower. Thus each insecticide reduces the selection pressure for resistance to the other and could be said to 'protect' the other. This protection means that evolution of resistance to one insecticide is always slower when it is used in a mixture than when used alone. This protection has been termed multiple intra-generational killing [@Consortium2013] because individuals have the potential to be killed by more than one insecticide.

Whilst this protection ensures that resistance to each insecticide evolves more slowly in a mixture it does not guarantee that a mixture strategy is preferable  to a sequence once the resistance to both insecticides is taken into account. In a mixture, selection pressure is applied by both insecticides at the same time, whereas in a sequence there is no selection for the insecticide not being used. Resistance can evolve more slowly for both insecticides in sequential use, despite being faster for each, because they occur one after the other. In Figure 5A it takes 60 generations for resistance to reach the threshold for each insecticide in the sequence, and 90 when they are applied together in the mixture, so the sequence is still favoured because 2x60 is greater than 90.

This advantage of sequence over mixtures can be removed by increasing the effectiveness of one of the insecticides. Increasing the effectiveness of one insecticide (Fig 5B) decreases the total time for the sequence from 120 to 90 and increases that for the mixture from 90 to 115 thus changing the order and favouring the mixture. The mechanism for this change can be seen by comparing the shape of the curves in Figs 5A & B. In the mixture, resistance to insecticide 1 (with the increased effectiveness) rises at a similar speed in both figures. However, the increased effectiveness of insecticide 1 causes resistance to insecticide 2 to increase more slowly initially in the mixture. Once insecticide 1 reaches resistance of around 50% at around 50 generations, the curve for insecticide 2 becomes steeper. Once the first insecticide reaches the resistance threshold of 50% it kills fewer individuals that are resistant to the second insecticide, it loses its protective effect and thus stops slowing the rise in resistance to the second insecticide. A similar effect is visible when the effectiveness of both insecticides are increased (Fig 6C). The identical resistance curves for each insecticide in the mixture are shallower than at the lower effectiveness (Fig 5A) because more individuals resistant to each insecticide are being killed by the other insecticide. In this case the 'protection' given to both insecticides by the other declines at the same rate so we don't see the change in slope seen in Fig 5B. In both cases increasing the effectiveness of insecticides increases their protective effect when used in mixtures and favours mixtures over sequences.

Fig 5B also demonstrates that when two insecticides have a different effectiveness in a mixture, other parameters being equal, resistance will be expected to increase faster to the more effective insecticide (the red curve in this case). The more effective insecticide prompts both a) higher selection pressure for its own resistance and b) greater 'protection' reducing the rise in resistance to the other insecticide. The more effective insecticide has a faster evolution of resistance than the less effective one but this is not caused by the presence of the less effective one. Indeed the presence of the less effective one is still slowing the evolution of resistance to the more effective one, as indicated by the red solid curve for use in the mixture being shallower than the red dashed curve for use alone as part of the sequence.

#### Difference between exposure and effectiveness
It seems initially counter-intuitive that whereas increasing effectiveness slows the development of resistance for mixtures, increasing exposure speeds it up. Increasing the exposure to both insecticides results in a decrease in time-to-resistance (from 90 generations in Fig 5A to 40 generations in Fig 5C) where increasing the effectiveness of both insecticides by the same amount results in a slight increase (from 90 in Fig 5A to 115 in Fig 5B). This contrasts with the identical effects that exposure and effectiveness have on a single insecticide (Figs 3 A,B) and on insecticides in sequence (the red dashed lines are the same in Figs 5B & 5C). This difference, that exposure reduces times-to-resistance for a mixture and effectiveness increases it, was unexpected for us and we are keen to explore it in more detail. The most likely explanation is that increasing exposure increases overall selection pressure and lowers time to resistance, whereas increasing effectiveness increases mutual protection of the insecticides in a mixture and lowers the speed at which resistance evolves.

#### Costs of resistance
Fitness costs of resistance have been widely reported(review in [@Kliot2012]) but it has been suggested that few of these are directly relevant to control failure in the field [@Ffrench-Constant2017]. Because of this likely rarity of costs we did not include them in most of our analyses (with the exception of Fig 9).

Costs of resistance favour mixtures in these analyses because the evolution of resistance is slowed more by cost in a mixture than in a sequence (compare the red solid lines for a mixture to the red dotted line for a sequence in Fig 9C and D). It makes sense that the protective effect of the other insecticide in a mixture combines with the cost of resistance to reduce selection pressure more than in a sequence.

The resistance curves for two insecticides in a mixture that differ only in their cost of resistance, diverge more as the frequency of resistance increases (Fig 9C). This makes intuitive sense, as the frequency of resistance genes increases the impact of fitness costs associated with those genes also increases. Thus differences due to cost are less noticeable at low resistance frequencies and become more noticeable at higher resistance frequencies. Thus if a lower resistance threshold was used cost would have less of an effect and if a higher resistance threshold was used cost would have more of an effect. Also note that if the costs are set even higher than we set them here (e.g. above 0.2) they can maintain resistance frequencies below 1 or lead to their decline. However, we do not expect such high costs to occur in operational settings.

@@

Earlier modelling work has made sometimes contrasting predictions and these can be explained by the details of their structure and input choices. One model predicted that "the use of mixtures is always more effective in delaying the onset of resistance often by many orders of magnitude"[@Mani1985]. That model assumed that SS genotypes were always killed, RR always survived, and that a constant 10% of mosquitoes escaped the insecticide. This would be represented in our model by a very high 'insecticide effectiveness' of 1, a 'resistance restoration' of 1 and an 'exposure' of 0.9. This result is consistent with our prediction that the very high effectiveness is likely to favour mixtures over sequences (e.g. see Fig 6D and Table 4).

@@

#### Poor implementation
One aspect of the results presented here can seem to contradict published advice and has been questioned when we have presented model results. Our model predicts that lower insecticide effectiveness or exposure will lead to slower spread of resistance in sole use and sequences (Figs 3 & 5, Table 2). This is consistent with expectations that lower effectiveness and exposure would lead to lower mortality of susceptibles and thus lower selection pressure for resistance [e.g. @Roush1998]. This can seem to contradict recommendations that poor implementation of insecticide interventions can promote the development of resistance. e.g. from [@FAO2012] "Certain pest control practices have consistently been shown to exacerbate the loss of susceptible pest populations and the development of resistance. These include: ... - the use of application rates that are below or above those recommended on the label, - poor coverage of the area being treated ...".

Poor implementation would likely cause lower effectiveness of, and exposure to, insecticides, so how can this increase the potential for resistance when our model suggests it would be decreased ? The explanation for this apparent difference is in the effect of dominance. Poor application can reduce the mortality of heterozygotes (SR) thus effectively increasing the dominance of the resistant allele and increasing its rate of evolution [@Bourguet2000][@Barbosa2011][@FAO2012] see Fig 1 of [@Levick2017]. Our model results are consistent with this, showing that increasing dominance of restoration leads to faster spread of resistance (Fig 3C, 7). So, the effect of poor implementation of an insecticide intervention on resistance could be either positive or negative depending on the relative effects on dominance of the resistance gene, insecticide effectiveness and exposure. Similarly, the effects of poor application on the relative benefits of mixtures and sequences will depend on these trade-offs because of our result that higher effectiveness favours mixtures, higher exposure favours sequences and higher dominance favours neither (Table 4). More research into how the mortality of different genotypes (susceptible, resistant and heterozygous) is affected by poor application rates or insecticide degradation over time would help make this clearer.

@@

In addition we provide an on-line user interface to the model (Fig 2) [@South2017a] enabling readers with no coding experience to investigate model behaviour in areas of parameter space we have not shown. The user-interface allows inputs to be set for two scenarios and results to be viewed side by side. For example you can set a higher insecticide effectivenesses in one scenario to see how this favours mixtures over sequences (Fig 2) or increase the exposure levels to see how this favours sequences.

>>>>

```{r, eval=TRUE, include=FALSE}
#!!!! PAPER PROTOCOL
#!!!! submission MS, knit to word_document 
#!!!!  in WORD: delete figs, set double space, add line numbers 
#!!!!  (spacing & line numbers work for pdf output but not for word)
library(knitr)
opts_chunk$set(dev="tiff",
               dev.args=list(compression="lzw"),
               dpi=300,
               cache=FALSE,
               fig.path='pap2figs/')
```

@@

The calculation of fitness for each genotype when considering just a single insecticide is illustrated in Figure 2. In the model itself this is repeated for both insecticides and the fitnesses multiplied. Firstly the exposure input determines the proportion of the population in the left and right panels (exposed and not exposed). For those that are exposed (left panel) insecticide effectiveness sets the fitness for SS and resistance restoration 'restores' a portion of the fitness for RR. A value of 1 for resistance restoration would bring the fitness of RR back to 1, equivalent to the unexposed SS and meaning the RR genotype is completely unaffected by the insecticide. A value of 0 for resistance restoration would lead to the RR having the same fitness as SS in the presence of the insecticide and effectively there would be no resistance. We use 'Resistance restoration' to represent how much the RR genotype restores fitness in the presence of the insecticide. Cost of resistance determines the fitness of RR in the portion of the population not exposed to the insecticide.

Dominance of restoration determines how the fitness for SR sits between that of SS and RR for those that are exposed to the insecticide. For those that are not exposed, fitness of SS is set to 1 by definition and resistance cost determines the fitness of RR. Dominance of cost determines how the fitness for SR sits between that of SS and RR in the absence of the insecticide (Fig 2).

The simulation proceeds through non-overlapping generations. In each generation selection is represented by multiplying genotype frequencies in the population by their relative fitness, the latter depending on their absolute fitness (Fig 2) and level of exposure. This acts to make the fitter alleles more common over time. Separate sexes and standard sexual reproduction with recombination is included. There is also the potential to set exposure rates to be different for males and females to represent their different behaviours but we do not explore that parameter here.

@@

A major aim of this paper is to us modern techniques to make the modelling accessible to a wider community. The model is implemented in R [@RCoreTeam2016], the code is open-source and hosted on Github [@South2017] including the code to generate both the figures and text of this paper. We provide an on-line user interface [@South2017a] (Fig 3) accessing the same code and enabling readers with no coding experience to change inputs and run the model themselves.

@@

### Single insecticide
For single insecticide use, higher values of insecticide effectiveness, exposure, resistance restoration and its dominance all resulted in faster resistance spread (Fig 4 A-D) and thus fewer generations to reach a resistance threshold of 0.5. This makes intuitive sense as all increase the impact of the insecticide and thus increase selection for resistance. Similarly, higher values of starting resistance frequency also resulted in shorter times to resistance thresholds (Fig 5A) through the even simpler mechanism that a smaller change in frequency is required to reach the threshold.

Cost of resistance (Fig 5B) and its dominance (not shown) had the opposite effect. Higher values led to a slower development of resistance. Again this is intuitive because costs reduce the advantage of resistance and decrease selection. These effects of inputs on single insecticide use are summarised in Table 2.

@@

### Two insecticides
When two insecticides are used we can compare the relative performance of concurrent use in a mixture with sequential use. Each panel in Figures 6-10 compares mixture and sequence for a single combination of inputs. To help understand the mechanisms influencing this relative performance we start with a single base scenario and demonstrate changing inputs individually. In this base scenario we set effectiveness, exposure, resistance restoration, and dominance at 0.5, the starting frequencies of resistance to 0.01 and cost of resistance to 0. Resistance arises slower for sequential than mixture for this base scenario (Fig 6A). Resistance to both insecticides in the mixture follow the same path and reach the threshold at the same time (as would be expected given that they have identical input parameters). When used in sequence the curve (dashed) for each insecticide individually is steeper than in the mixture but, because they happen one after the other, it takes longer for both to reach the resistance threshold (Fig 6A). Thus the time-to-resistance for the mixture divided by the sequence is less than 1 at 0.8.

#### Insecticide exposure and effectiveness
Increasing exposure (the proportion of insects that come into contact with the insecticides) (Fig 5C) decreases the time-to-resistance for both the sequential and mixture strategy. The effect on the mixture is greater so time-to-resistance remains longer for the sequence (the ratio decreasing from 0.8 to 0.6). In contrast keeping the exposure constant and increasing effectiveness (the proportion of SS insects that are killed by contact) of one of the insecticides (Fig 6B) results in a longer time-to-resistance for the mixture relative to the sequence (and the ratio increasing from 0.8 to 1.2). The longer time-to-resistance for the mixture results from a slower increase in resistance for the less effective insecticide. Resistance to the less effective insecticide in the mixture increases slowly initially and then speeds up after the more effective insecticide has reached the resistance threshold. This points a mechanism whereby the more effective insecticide initially slows the rate of evolution of resistance to the less effective one.

The result is that a more effective insecticide increases time-to-resistance when used in a mixture (compare the solid lines in Fig 6A & B). This is opposite to what happens when used in sequence (compare the red dashed line in Fig 6A & B) and in isolation (Fig 4A), when more effective insecticides shortened times-to-resistance.

In summary increasing effectiveness can favour mixtures moving from column 1 to 2 in Figure 6, and increasing exposure can favour sequences moving from row 1 to 2 in Figure 6. Reduced male exposure with respect to female, as might be expected for male mosquitoes not seeking blood meals, had a similar effect to changing overall exposure. Reducing male exposure (not shown) increased times-to-resistance and favoured mixtures over sequences.

Increasing the effectiveness further of either or both insecticides increases the performance of mixtures over sequences in terms of time-to-resistance. Figure 7A uses the same inputs as Fig 6B. From this scenario a higher effectiveness for insecticide2 (Fig 7C) or insecticide 1 (Fig 7B) or both (Fig 7D) all result in a greater positive difference in time-to-resistance for mixture over sequence. This again points to how more effective insecticides can slow resistance to another when used in a mixture.

#### Resistance restoration and dominance
Increasing resistance restoration or its dominance decreases time-to-resistance for both sequences and mixtures (Fig 8) in the same way that it did for sole use (Fig 4). The result is that increasing either or both of dominance and resistance restoration (Figs 8B,C,D) doesn't change the relative time-to-resistance for mixtures and sequences from that in the base scenario (Fig 8A) where it takes resistance longer to develop for the sequence. So, unlike effectiveness and exposure, the levels of resistance restoration and dominance do not effect whether a mixture or a sequence are likely to be favoured.

#### Starting frequencies of resistance
Changing the starting frequency of resistance had a similar effect on time to resistance for both sequences and mixtures and thus had little effect on their relative performance. For example taking the base scenario and reducing the starting frequency of one resistance allele did not change from sequence being favoured (compare Fig 9C to 9A). Similarly taking a scenario in which time-to-resistance is longer for a mixture (Fig 9B) and decreasing the starting frequency of that resistance allele did not change the fact that the mixture was favoured (Fig 9D).

#### Costs of resistance
Increasing the cost of resistance to one insecticide in a mixture leads resistance to that insecticide to evolve more slowly (compare the solid red line to the solid blue line in Figure 10C, the red line is for the resistance with the higher cost). The resistance with the higher cost also increases more slowly in the mixture relative to when it is used in sequence (compare the solid red line to the dotted red line in Fig 10C). The greater effect of cost on mixtures means the advantage of sequences over mixtures in Fig 10A (and Fig 6A) is removed in Fig 10C. With costs set to 0 time-to-resistance for mixture divided by sequence is 0.8 in Fig 10A and 1 in Fig 10C. At higher insecticide effectiveness the benefit of mixture over sequence is improved as seen by comparing Fig 10D to 10B.

@@

#### Linkage disequilibrium
It has previously been suggested [@Curtis1985][@WHO2012] that using an insecticide to which there was existing resistance in a mixture with a new insecticide with little resistance could lead to more rapid evolution of resistance to the latter through a process termed linkage disequilibrium. Figure 9 C&D show how our model predicts that linkage disequilibrium does not lead to more rapid evolution of resistance to the newer insecticide. In Figures 9 C&D the newer insecticide with lower starting resistance (in red) leads to faster resistance when used alone (red dotted) than when used in a mixture (red solid).

@@

When single insecticide use was investigated (Figs 4-5) higher values of parameters that increased the selective advantage of resistance all led to more rapid evolution resistance, whereas higher values for cost led to slower evolution of resistance (Fig 5B). Mechanisms explaining the evolution of resistance to single insecticide use, are summarised in Table 2. Referring to Figure 2 can help explain these responses. Exposure and effectiveness act by decreasing the fitness of SS and SR more than RR, dominance of restoration acts by increasing fitness of SR and resistance-restoration acts by increasing fitness of RR & SR. Cost acts by decreasing the fitness of RR and SR among those mosquitoes  not exposed.

#### Two insecticides
Responses to input parameters are different when two insecticides are used either in sequence or in mixtures as compared to when used alone. These differences can help us to understand the dynamics of resistance evolution. In each panel of Figures 6-10 the solid lines, indicating that an insecticide is being used at the same time as another in a mixture, are always shallower than the same colour dashed line indicating the insecticide is being used in isolation as part of a sequence. The shallower curves for the mixture can be explained by each insecticide killing individuals that are resistant to the other insecticide. Because resistant individuals are being killed, the selection pressure for that resistance is lower. Thus each insecticide reduces the selection pressure for resistance to the other and could be said to 'protect' the other. This protection means that evolution of resistance to one insecticide is always slower when it is used in a mixture than when used alone. This protection has been termed multiple intra-generational killing [@Consortium2013] because individuals have the potential to be killed by more than one insecticide.

Whilst this protection ensures that resistance to each insecticide evolves more slowly in a mixture it does not guarantee that a mixture strategy is preferable  to a sequence once the resistance to both insecticides is taken into account. In a mixture, selection pressure is applied by both insecticides at the same time, whereas in a sequence there is no selection for the insecticide not being used. Resistance can evolve more slowly for both insecticides in sequential use, despite being faster for each, because they occur one after the other. In Figure 6A it takes 60 generations for resistance to reach the threshold for each insecticide in the sequence, and 90 when they are applied together in the mixture, so the sequence is still favoured because 2x60 is greater than 90.

This advantage of sequence over mixtures can be removed by increasing the effectiveness of one of the insecticides. Increasing the effectiveness of one insecticide (Fig 6B) decreases the total time for the sequence from 120 to 90 and increases that for the mixture from 90 to 115 thus changing the order and favouring the mixture. The mechanism for this change can be seen by comparing the shape of the curves in Figs 6A & B. In the mixture, resistance to insecticide 1 (with the increased effectiveness) rises at a similar speed in both figures. However, the increased effectiveness of insecticide 1 causes resistance to insecticide 2 to increase more slowly initially in the mixture. Once insecticide 1 reaches resistance of around 50% at around 50 generations, the curve for insecticide 2 becomes steeper. Once the first insecticide reaches the resistance threshold of 50% it kills fewer individuals that are resistant to the second insecticide, it loses its protective effect and thus stops slowing the rise in resistance to the second insecticide. A similar effect is visible when the effectiveness of both insecticides are increased (Fig 7C). The identical resistance curves for each insecticide in the mixture are shallower than at the lower effectiveness (Fig 6A) because more individuals resistant to each insecticide are being killed by the other insecticide. In this case the 'protection' given to both insecticides by the other declines at the same rate so we don't see the change in slope seen in Fig 6B. In both cases increasing the effectiveness of insecticides increases their protective effect when used in mixtures and favours mixtures over sequences.

Fig 6B also demonstrates that when two insecticides have a different effectiveness in a mixture, other parameters being equal, resistance will be expected to increase faster to the more effective insecticide (the red curve in this case). The more effective insecticide prompts both a) higher selection pressure for its own resistance and b) greater 'protection' reducing the rise in resistance to the other insecticide. The more effective insecticide has a faster evolution of resistance than the less effective one but this is not caused by the presence of the less effective one. Indeed the presence of the less effective one is still slowing the evolution of resistance to the more effective one, as indicated by the red solid curve for use in the mixture being shallower than the red dashed curve for use alone as part of the sequence.

#### Difference between exposure and effectiveness
It seems initially counter-intuitive that whereas increasing effectiveness slows the development of resistance for mixtures, increasing exposure speeds it up. Increasing the exposure to both insecticides results in a decrease in time-to-resistance (from 90 generations in Fig 6A to 40 generations in Fig 6C) where increasing the effectiveness of both insecticides by the same amount results in a slight increase (from 90 in Fig 6A to 115 in Fig 6B). This contrasts with the identical effects that exposure and effectiveness have on a single insecticide (Figs 4 A,B) and on insecticides in sequence (the red dashed lines are the same in Figs 6B & C). This difference, that exposure reduces times-to-resistance for a mixture and effectiveness increases it, was unexpected for us and we are keen to explore it in more detail. The most likely explanation is that increasing exposure increases overall selection pressure and lowers time to resistance, whereas increasing effectiveness increases mutual protection of the insecticides in a mixture and lowers the speed at which resistance evolves.

#### Costs of resistance
Fitness costs of resistance have been widely reported(review in [@Kliot2012]) but it has been suggested that few of these are directly relevant to control failure in the field [@Ffrench-Constant2017]. Because of this likely rarity of costs we did not include them in most of our analyses (with the exception of Fig 10).

Costs of resistance favour mixtures in these analyses because the evolution of resistance is slowed more by cost in a mixture than in a sequence (compare the red solid lines for a mixture to the red dotted line for a sequence in Fig 10C and D). It makes sense that the protective effect of the other insecticide in a mixture combines with the cost of resistance to reduce selection pressure more than in a sequence.

The resistance curves for two insecticides in a mixture that differ only in their cost of resistance, diverge more as the frequency of resistance increases (Fig 10C). This makes intuitive sense, as the frequency of resistance genes increases the impact of fitness costs associated with those genes also increases. Thus differences due to cost are less noticeable at low resistance frequencies and become more noticeable at higher resistance frequencies. Thus if a lower resistance threshold was used cost would have less of an effect and if a higher resistance threshold was used cost would have more of an effect. Also note that if the costs are set even higher than we set them here (e.g. above 0.2) they can maintain resistance frequencies below 1 or lead to their decline. However, we do not expect such high costs to occur in operational settings.

@@

Earlier modelling work has made sometimes contrasting predictions and these can be explained by the details of their structure and input choices. One model predicted that "the use of mixtures is always more effective in delaying the onset of resistance often by many orders of magnitude"[@Mani1985]. That model assumed that SS genotypes were always killed, RR always survived, and that a constant 10% of mosquitoes escaped the insecticide. This would be represented in our model by a very high 'insecticide effectiveness' of 1, a 'resistance restoration' of 1 and an 'exposure' of 0.9. This result is consistent with our prediction that the very high effectiveness is likely to favour mixtures over sequences (e.g. see Fig 7D and Table 4).

@@

#### Poor implementation
One aspect of the results presented here can seem to contradict published advice and has been questioned when we have presented model results. Our model predicts that lower insecticide effectiveness or exposure will lead to slower spread of resistance in sole use and sequences (Figs 4 & 6, Table 2). This is consistent with expectations that lower effectiveness and exposure would lead to lower mortality of susceptibles and thus lower selection pressure for resistance [e.g. @Roush1998]. This can seem to contradict recommendations that poor implementation of insecticide interventions can promote the development of resistance. e.g. from [@FAO2012] "Certain pest control practices have consistently been shown to exacerbate the loss of susceptible pest populations and the development of resistance. These include: ... - the use of application rates that are below or above those recommended on the label, - poor coverage of the area being treated ...".

Poor implementation would likely cause lower effectiveness of, and exposure to, insecticides, so how can this increase the potential for resistance when our model suggests it would be decreased ? The explanation for this apparent difference is in the effect of dominance. Poor application can reduce the mortality of heterozygotes (SR) thus effectively increasing the dominance of the resistant allele and increasing its rate of evolution [@Bourguet2000][@Barbosa2011][@FAO2012] see Fig 1 of [@Levick2017]. Our model results are consistent with this, showing that increasing dominance of restoration leads to faster spread of resistance (Fig 4C, 8). So, the effect of poor implementation of an insecticide intervention on resistance could be either positive or negative depending on the relative effects on dominance of the resistance gene, insecticide effectiveness and exposure. Similarly, the effects of poor application on the relative benefits of mixtures and sequences will depend on these trade-offs because of our result that higher effectiveness favours mixtures, higher exposure favours sequences and higher dominance favours neither (Table 4). More research into how the mortality of different genotypes (susceptible, resistant and heterozygous) is affected by poor application rates or insecticide degradation over time would help make this clearer.

@@

In addition we provide an on-line user interface to the model (Fig 3) [@South2017a] enabling readers with no coding experience to investigate model behaviour in areas of parameter space we have not shown. The user-interface allows inputs to be set for two scenarios and results to be viewed side by side. For example you can set a higher insecticide effectivenesses in one scenario to see how this favours mixtures over sequences (Fig 3) or increase the exposure levels to see how this favours sequences.
